When the Pen Becomes a Sword: Retired US Military Officers and Op-Ed Use in Contemporary America

### 2020-07-13 code/Submissionv1.Rmd
commit changes

  _This paper explores variation in the professionalism of opinion pieces authored by retired US military officers and published in major US newspapers in the past 40 years. Assembling and analyzing more than 330 opinion-editorial pieces published by retired military officers in The Wall Street Journal, The Washington Post, The New York Times, The Los Angeles Times, and the USA Today since 1979 reveals that there has been an increasing trend in published opinion pieces which fail to uphold professional standards as defined by military norms. The paper further argues that this trend is primarily the result of two significant strains on the military's politically neutral culture: political polarization and military credibility. Finally, the paper develops a theory that shows how the forces of political polarization and military credibility interact, and argues that the norm of a professional, apolitical military will remain under further strain in the current US political environment._  
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: researchpaper-ps904.bib # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs

@@

In late May and early June 2020, the American public was gripped by escalating racial tension and discussions about police brutality following the death of George Floyd, an unarmed black man, by police in Minneapolis. Protests, most of them peaceful, were accompanied in some cases by rioting and looting. On June 1, after giving remarks at the Rose Garden, President Donald Trump walked to St. John's Church after security forces visibly displaced throngs of protesters in neighboring Lafayette Square.  Flanked by members of his cabinet, to include the Secretary of Defense, and with the Chairman of the Joint Chiefs of Staff not far behind, the President held up a Bible at the church, stared into the cameras of the media, and, several minutes later, returned to the White House.

@@

Retired military officers were among the elites who weighed in to give their opinions not only on the Trump administration's handling of specific events near the White House, but also on the broader issues of race in America. But for those paying attention to what these retired officers said, it is also clear that there was substantial variation in the tone and tenor of each officer's remarks. While all of them expressed deep concern about the possibility of active-duty forces confronting peaceful protesters in clear violation of the First Amendment of the US Constitution, [@brooks_dismay_2020; @dempsey_former_2020], some officers went much further, leveling direct criticisms of the President's fitness to be President [@goldberg_james_2020], his administration's broader policies on race [@allen_moment_2020], and his general leadership abilities [@mullen_i_2020].

This paper examines the way in which retired military officers have attempted to influence public opinion in contemporary America through opinion-editorial publications. It is an interesting question worthy of examination because, as I argue, at least three important factors are interacting in unique ways to shape how and why retired military officers speak out on a variety of issues. These factors are professionalism, polarization, and credibility. In particular, polarization and credibility are creating tremendous strains on the military's professional nature, and in some cases, are causing words to become weaponized.

This paper proceeds in four parts. First, I briefly review the concepts of military professionalism, polarization, and credibility in order to orient the reader to the theory that I build in part two. In the second section, I build a theory in which I argue that the contemporary forces of polarization and military credibility can strain the concept of professionalism such that words and actions become weaponized to the point of being antithetical to the very concept of professionalism itself. Third, I turn to the data, which consists of more than 330 opinion-editorial pieces published in major US newspapers by retired military officers since the late 1970s. While the data shows that the vast majority of these pieces are in fact in line with the notion of military professionalism, recent trends show that the retired military officers are weighing in on topics that fall outside of the realm of national security as well as expressing clear partisan preferences in their arguments. Fourth, I briefly explore the implications of these results on the military as an institution, and specifically, its ability to remain professional in a setting where polarization is high and the institution commands wide public support.

\singlespace
# Military Professionalism
\doublespace
The ideal of the US military acting as an objective, apolitical and non-partisan institution is deep. One of the strongest origins of the norm stems from the work of the political scientist, Samuel Huntington, who, in his 1957 work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, asserted that two distinct forces would always be in tension with one another in a liberal democracy such as the United States. The first was a "functional imperative," which Huntington said drove the requirement that a military be strong enough and capable of defeating the military threats it faced, and the second was a "societal imperative," which Huntington said generated requirements for the military in that the military could never lose so many aspects of it such that it no longer represented the liberal democracy that it defended [@huntington_soldier_1957, 1-3]. Huntington's point was that America needed a strong military that could win on the battlefield, but cautioned that such a military could not be so foreign in character so as to alienate the tenets of an American liberal society. The key to resolving the tension generated by these two opposing forces was to adopt a scheme of "objective civilian control" in which political leaders established the ends and goals of policy, while military leaders developed and implemented the military plans to accomplish these goals [@huntington_soldier_1957, 80-85]. To Huntington, largely separate operating spheres between the political leaders of a democracy and the military as an institution would ensure that the military remained "professional," and thus, untainted by partisan interests and concerns. In this way, Huntington posited that the American military would best meet the task of defeating security threats while also remaining apolitical.

While Huntington's theory has enjoyed tremendous staying power (Huntington is still studied and taught in many American military institutions), he is not without critics [see, for example, @feaver_civil-military_1996; @cohen_supreme_2003; @feaver_armed_2005; and @brooks_paradoxes_2020]. It would be too burdensome to unpack all of those here, but one of the earliest and richest critiques of Huntington, and perhaps one of the most relevant given the potential influences of polarization and credibility, came from S.E. Finer. Authored in 1962, Finer's book, _The Man on Horseback: The Role of the Military in Politics_, argued that Huntington's concept of professionalism simply was not strong enough to keep the military out of politics [@finer_man_1962]. Finer's concern was that of military officers adopting an ethic of professionalism such that they would actually be more likely - not less likely - to become involved in political issues [@finer_man_1962, 22-23]. Finer was especially concerned that the adoption of professionalism alone would lead to long-serving military professionals believing that they knew what was best for their country over and above that of temporary elected civilian leaders [@finer_man_1962, 22-23] Finer warned that "the moment the military draw this distinction between nation and the government in power, they begin to invent their own private notion of the national interest, and from this it is only a skip to the constrained substitution of this view for that of the civilian government…this purported care for the national interest as defined by the military is indeed one of their main reasons for intervening [@finer_man_1962]. Finer's suggestion was that because of the perils of professionalism as espoused by Huntington, what was really needed to keep the military out of politics was a deep internalization of "the principle of civil supremacy" [@finer_man_1962]^[
Importantly, neither Huntington's concept of professionalism nor Finer's critique of the same are arguing for blind obedience by the military to civilian orders and directions, a point that is, in the author's opinion, often missed in debates about the perils of professionalism. On the contrary, both Huntington and Finer make room and acknowledge that there are instances when military disobedience is not only required, but justified. However, both agree that in general terms, these situations should be exceedingly rare.]  
It is not only in the academic literature that the tradition and worth of an apolitical military has been cultivated. The operational military has also been central in upholding the military's non-partisan traditions, albeit with a far less than perfect record. Retired Army General Douglas MacArthur, a hero of World War Two in the Pacific and who later commanded the occupation of Japan, is well known for his disagreement with President Truman regarding American strategic goals during the Koran War. Yet his remarks to cadets at West Point as an 82 year old retired officer are also well known and applicable to the topic at hand:

@@

Current military doctrine and practices likewise reinforce the notion of the military institution remaining apolitical. Prior to the 2016 Presidential Election, Marine Corps General Joseph Dunford, who then served as the Chairman of the Joint Chiefs of Staff, said that the basis upon which the military is able to maintain the public's trust is it's ability to remain "apolitical and neutral" [@dunford_upholding_2016]. Retired General Martin Dempsey, in an opinion piece published in _The Washington Post_ on July 31, 2016, specifically criticized retired Army Major General Mike Flynn and Marine Corps General John Allen for participating in the Republican and Democratic conventions, respectively, stating that, "As generals, they have an obligation to uphold our apolitical traditions. They have just made the task of their successors – who continue to serve in uniform and are accountable for our security - more complicated. It was a mistake for them to participate as they did. It was a mistake for our presidential candidates to ask them to do so" [@dempsey_military_2016].

Although the notion of an apolitical military is the theoretical goal of the institution, the practical record remains far from perfect. From MacArthur's clash with President Truman in the early 1950's, to the so-called "Revolt of the Generals" in 2005-2006, military leaders, even those who are retired, have ventured into political territory from time to time. What we have established, though, is clear: a non-partisan military that performs its military mission well has long been the goal of both civil-military relations theorists and military officers. More importantly, this is what fosters and ultimately earns the trust and confidence of the American people.

@@

Military credibility remains high among the US population, though this is not something that has come easy for the military. Consider that in the late 1970s, as a likely result of public opinion over the Vietnam War, 54% of Americans indicated that they trusted the military "a great deal or quite a lot," whereas in 2019, 73% of Americas expressed the same sentiments [@gallup_polls_confidence_2007]. Additionally, military credibility is high relative to other institutions. According to a recent Pew Research Center poll, 80% of the American public expects the military to act in the best interests of the American public, compared to business leaders (45%), the media (40%), and elected officials (25%) [@johnson_trust_2018].

When credibility between the American population and the military as in institution is high, it can be inferred that the public trusts the military - to accomplish its military mission most of all, but also to do more than that - to develop healthy institutional cultures, to vigorously address unethical conduct, and to exercise good leadership while taking care of American service members [@us_army_repairing_2014]. Senior military leaders are aware of the importance of maintaining public trust, and are generally sensitive to issues or concerns which may lead to a loss in confidence of the military [@us_army_repairing_2014].

@@

## Political Polarization

Although scholars have offered differing explanations for the causes of polarization, ranging from concerns over race and religion [@abramowitz_great_2018], to the size and role of government [@sides_two_2015], to fierce policy disagreements over policy issues [@mann_its_2012], virtually all scholars agree that polarization in America has become more pronounced in recent decades [@hare_polarization_2014; @bonica_mapping_2014], and by many measures, is worse than it has been during any period since the end of the American Civil War [@hare_polarization_2014]. Scholars have also shown that whereas polarization in previous decades was pronounced over singular and specific issue areas, to include redistribution policies following key Civil Rights legislation passed in the 1960s [@hare_polarization_2014], recent polarization is marked by a phenomenon known as "conflict extension," [@layman_party_2002], referring to the notion that a far wider set of issues and policy preferences covering the full gamut of social welfare, racial, and cultural questions are taking on a polarizing character [@layman_party_2006].

The 2016 Presidential Election, resulting in the Trump presidency, confirms the nature and extent of polarized politics in America. Political scientists, journalists, and reporters are still trying to piece together the specific factors that led to Donald Trump's surprise victory. Sides, Tesler, and Vavreck argue that Trump's election reveals just how split Americans are, particularly on core issues of identity and attitudes regarding race and religion [@sides_identity_2018]. During the years of the Trump Presidency, moreover, the American people have indicated that polarization is an issue whose presence is being felt and worsening [@schaeffer_far_2020]. Finally, given that Americans feel more strongly today about more issues and especially those issues from which individuals draw a sense of core identity, it is hardly surprising that individuals seem to apply partisan lenses, or biases, to what they hear. Michael Robinson showed as a result of political polarization, partisan groups are cognitively biased in how they receive and process information [@robinson_who_2019]. As a result, Robinson found that partisans will express confidence in the military in vastly differing ways even after neutral, non-partisan information about various types of military failure is presented to these groups [@robinson_who_2019].

## When Words Become Weaponized - Theory

The purpose of this section is to explain how the concepts of professionalism, credibility, and political polarization can interface to create a potentially dangerous atmosphere for both military professionals and the public. It must be stressed from the outset that none of these concepts in isolation impose a normative prohibition on the act of a military officer taking up the pen to write an opinion piece. Indeed, Huntington's concept of military professionalism recognizes the unique skills of military officers, and Zaller's framework of elite discourse tacitly affirms that retired military officers, especially generals and admirals, most of whom possess more than 30 and in some cases 40 years of professional experience, possess relevant expertise that the public would find helpful, relevant, and convincing as it pertains to a host of potential policy debates. Even the existence of political polarization itself is insufficient to suggest that retired military officers should never give their opinion on specific matters.

One can reasonably infer that a professional opinion piece written by a military officer would then take on three very important qualities. First, stemming from the concept of military professionalism, the piece would address a relevant military topic. To be clear, multistar generals and admirals often are experienced in the realm of strategy, a word that is often misunderstood and misapplied in discussions of the topic [see for example @strachan_direction_2014]. Without unpacking the concept of strategy further, the relevant point is that military elites, and especially those who achieved a high rank, have been privy to decisions and experiences occurring near the highest levels of government. Their body of experiences is not confined to mere tactics. They They are familiar with issues ranging from how wars should be best fought, to future security threats looming on the horizon, to fixing the institutional culture of a military service. Second, from a credibility and expertise perspective, a professional piece would be based on the officer's personal experiences and would be written from an individual's perspective so as not to infer that an argument is being made on behalf of the collective service or element within the military body. Third, an opinion made by a military officer would be free of any hint of partisan sentiment, whether advocacy or denouncement. If one of these three qualities - addressing a relevant military topic, from the perspective of an individual, and free of partisan expression - fails to be upheld, there is increasing likelihood that the officer's words can serve as a weapon, and, like other kinds of weapons, can do great harm.

Figure 1, below, outlines the general theory and framework of the impact of political polarization and institutional credibility on military professionalism, and subsequently, on various military outcomes. The system of variables represented in Figure 1 constitute the "drive chain" that can lead to the words of retired military officers becoming militarized.

When military credibility is high, a demand signal is created by the American public such that future interactions between the military institution and the American public are encouraged and welcomed. From an institutional perspective, the military might respond to the public's demand signal for increased military interaction by increasingly participating in public ceremonies, parades, and other showcases of military prowess (at sporting venues, for instance). From an individual perspective, military elites, especially retired military officers, would be psychologically primed to increasingly dialogue with the American public when military credibility is high. By extension, other outlets such as the Press would also be primed to showcase military elites when military credibility runs high, because these elites represent an institution that commands a high degree of public respect.

When military credibility is low, I theorize the opposite will occur. Since demand for interactions between the American public and the military is muted when institutional credibility for the military is low, the institution will not engage with the public as frequently nor as intimately because the strong demand signal that existed when institutional credibility was high no longer exists. At the individual perspective, military elites will also not engage, nor will the Press ask these elites to, with the American public. In short, institutional credibility and respect for the military institution impacts the likelihood of the military institution and military elites interacting with the American public. Such interaction can take a variety of forms. When respect for the military is high, the military is likely to interact, speak to, and dialogue with the American public, and when institutional credibility is low, the military is less likely to do so.

```{r, include=FALSE, echo=FALSE}
problem_dag <- dagify(
    Y ~ X1,
    X1 ~ X2 + X3,
    X2~X4,
    X3~X4,
    coords = tribble(
      ~ name,      ~ x,    ~ y,
      "Y", 1, .5,
      "X1", .7, .5,
      "X2", .3, 1,
      "X3", .3,0,
      "X4", 0, .5
    ))

@@

problem_dag %>% tidy_dagitty() %>% 
  dag_label(labels=c("Y"="Military\n Outcomes", 
                     "X1"="Military\n Professionalism", 
                     "X2"="Political\n Polarization", 
                     "X3"="Institutional\n Credibility", 
                     "X4"="Major Changes\n In American Society")) %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text()+ 
  geom_dag_label_repel(aes(label=label, fill=label), col="white", show.legend=FALSE)+ theme_dag() + labs(title="Figure 1", subtitle="The Impact of Political Polarization and Institutional Credibility on Military Professionalism")

```
The degree of political polarization evident also has the potential to impact the professionalism of the military as an institution. First, because of the phenomenon known as "conflict extension" [@layman_party_2002], which refers to the fact that the political parties have grown further apart from each other on an increasing number of issues, political polarization raises the number and scope of issues to disagree about. Research has consistently shown that the polarizing politics present in the United States today has caused voters who identify as liberals to be more liberal across multiple issue categories, and that the same holds for conservatives. The result is that in the past, the prospect of cooperation on issues that were not polarized remained high, whereas today, the prospect for cooperation in general is lower [@layman_activists_2010]. As it concerns the military institution, "conflict extension" [@layman_party_2002] increases the likelihood that members of the military feel strongly about an increasing number of issues. In other words, I contend that today's polarized environment makes it more difficult for an institution that to be politically neutral when there are fewer issues to remain neutral over. At the individual level, the same is true: as the non-partisan space seems to shrink, individuals will be enticed to not remain neutral.

>>>>

  _This paper explores variation in the professionalism of opinion pieces authored by retired US military officers and published in major US newspapers in the past 40 years. In doing so, this paper develops a theory that argues that two societal forces, political polarization and military credibility, interact in unique ways that ultimately strain the individual military officer's concepts of loyalty and duty in important ways. Assembling and analyzing more than 330 opinion-editorial pieces published by retired military officers in The Wall Street Journal, The Washington Post, The New York Times, The Los Angeles Times, and the USA Today since 1979, the paper argues there has been an increase in published opinion pieces which fail to uphold professional standards as defined by historical military norms. The paper concludes by arguing that today's political polarization reveals a profoundly changing society, and that the norm of a professional, apolitical military will remain under further strain in the current political environment._  
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: researchpaper-ps904.bib # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs

@@

In late May and early June 2020, racial tension and concerns about police brutality gripped the American public following the death of George Floyd, an unarmed black man, by police in Minneapolis. In several cities and communities across the nation, mostly peaceful protests were followed by riots and looting. On June 1, after giving remarks at the Rose Garden, President Donald Trump walked to St. John's Church after security forces visibly displaced throngs of protesters in neighboring Lafayette Square. Flanked by members of his cabinet, to include the Secretary of Defense, and with the Chairman of the Joint Chiefs of Staff not far behind, the President held up a Bible at the church, stared into the cameras of the media, and, several minutes later, returned to the White House.

@@

Additionally, several retired military officers were among the elites who weighed in to give their opinions not only on the Trump administration's handling of specific events near the White House, but also on the broader issues of race in America. But for those paying attention to what these retired officers said, it is also clear that there was substantial variation in the tone and tenor of each officer's remarks. While all of them expressed deep concern about the prospect of active-duty forces confronting peaceful protesters, and thus, violating protections held by the First Amendment of the US Constitution [@brooks_dismay_2020; @dempsey_former_2020], some officers went much further in their criticism, leveling direct insults of the President's fitness to lead [@goldberg_james_2020], his administration's broader policies on race [@allen_moment_2020], and his general leadership abilities [@mullen_i_2020].

This paper examines the way in which retired military officers have attempted to influence public opinion in contemporary America through opinion-editorial publications. It is an interesting question worthy of examination because, as I argue, at least three important factors are interacting in unique ways to shape how and why retired military officers speak out on a variety of issues. These factors are professionalism, political polarization, and military credibility. I argue that the latter two factors, political polarization and institutional credibility, are interacting in unique ways that create tremendous strains on the military's professional nature, and that these factors will continue to do so.

This paper proceeds in four parts. First, the paper briefly reviews the concept of military professionalism in order to orient the reader to the central role this concept has played in the ethos of the American military for the past several decades. Second, the paper builds a theory that demonstrates how the uniqueness of today's political polarization and institutional credibility are interacting to create an environment that threatens professional military behavior in ways that could be highly destructive. Third, the paper explores the data, which consists of more than 330 opinion-editorial pieces published in major US newspapers by retired military officers since the late 1970s. The analysis ultimately shows that although the vast majority of op-eds published by retired military officers maintain professional standards, there is clear evidence that retired military officers are using the pen to both argue about topics that fall outside of the realm of traditional national security topics, and to express clear partisan preferences when making their arguments. Fourth, the paper concludes by briefly exploring the implications of these results on the military as an institution, and specifically, on its ability to remain professional in a domestic setting marked by high political polarization in which the military commands wide public support.

\singlespace
# Military Professionalism as a Foundational Trait
\doublespace
The ideal of the US military acting as an objective, apolitical and non-partisan institution is storied and long. One of the strongest origins of the norm stems from the work of the political scientist, Samuel Huntington, who, in his 1957 work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, asserted that two distinct forces would always be in tension with one another in a liberal democracy such as the United States. The first was a "functional imperative," which Huntington said drove the requirement that a military be strong enough and capable of defeating the military threats it faced, and the second was a "societal imperative," which Huntington said generated requirements for the military in that the military could never lose so many aspects of it such that it no longer represented the liberal democracy that it defended [@huntington_soldier_1957, 1-3]. Huntington's point was that America needed a strong military that could win on the battlefield, but cautioned that such a military could not be so foreign in character so as to alienate the character of American liberal society. Huntington argued that the key to resolving the tension generated by these two opposing forces was to adopt a scheme of "objective civilian control" in which political leaders established the ends and goals of policy, while military leaders developed and implemented the military plans to accomplish these goals [@huntington_soldier_1957, 80-85]. To Huntington, largely separate operating spheres between the political leaders of a democracy and the military as an institution would ensure that the military remained "professional," and thus, untainted by partisan interests and concerns. In this way, Huntington posited that the American military would best meet the task of defeating security threats while also remaining apolitical.

While Huntington's theory has enjoyed tremendous staying power (Huntington is still studied and taught in many American military institutions), he is not without critics [see, for example, @feaver_civil-military_1996; @cohen_supreme_2003; @feaver_armed_2005; and @brooks_paradoxes_2020]. It would be too burdensome to unpack all of those here, but one of the earliest and richest critiques of Huntington, and perhaps one of the most relevant given the potential influences of political polarization and institutional credibility, came from S.E. Finer. Authored in 1962, Finer's book, _The Man on Horseback: The Role of the Military in Politics_, argued that Huntington's concept of professionalism simply was not strong enough to keep the military out of politics [@finer_man_1962]. Finer foresaw military officers adopting an ethic of professionalism such that they would actually be more likely, not less likely, to become involved in political issues [@finer_man_1962, 22-23].

Finer was especially concerned that the adoption of professionalism would lead to long-serving military professionals believing that they knew what was best for their country over and above that of temporary elected civilian leaders [@finer_man_1962, 22-23]. Finer warned that "the moment the military draw this distinction between nation and the government in power, they begin to invent their own private notion of the national interest, and from this it is only a skip to the constrained substitution of this view for that of the civilian government…this purported care for the national interest as defined by the military is indeed one of their main reasons for intervening" [@finer_man_1962]. In Finer's view, what was really needed to keep the military out of politics was not only the concept of professionalism, but rather a deep internalization of "the principle of civil supremacy" [@finer_man_1962].^[
Importantly, neither Huntington's concept of professionalism nor Finer's critique of the same are arguing for blind obedience by the military to all civilian orders and directions, a point that is, in the author's opinion, often missed in debates about military professionalism. On the contrary, both Huntington and Finer make room and acknowledge that there are instances when military disobedience is not only required, but justified. However, both agree that in general terms, these situations should be exceedingly rare.]

It is not only in the academic literature that the tradition and worth of an apolitical military has been cultivated. The operational military has also been central in upholding the military's non-partisan traditions, albeit with a far less-than-perfect record. Retired Army General Douglas MacArthur, a hero of World War Two in the Pacific and later, Commander of the occupation of Japan, is well known for his disagreement with President Truman regarding American strategic goals during the Korean War. And yet, the remarks MacArthur made to cadets at West Point in 1962, as an 82 year old retired officer, speak to the primacy of professional military conduct:

@@

Finally, current military doctrine and practices also reinforce the centrality of professional, apolitical, and non-partisan conduct. Prior to the 2016 Presidential Election, Marine Corps General Joseph Dunford, who at the time served as Chairman of the Joint Chiefs of Staff, said that the basis upon which the military is able to maintain the public's trust is it's ability to remain "apolitical and neutral" [@dunford_upholding_2016]. At nearly the same time, his predecessor, Retired Army General Martin Dempsey, published an opinion-editorial piece in _The Washington Post_ which criticized retired Army Major General Mike Flynn and Marine Corps General John Allen for participating in the Republican and Democratic conventions. Dempsey wrote, "As generals, they have an obligation to uphold our apolitical traditions. They have just made the task of their successors – who continue to serve in uniform and are accountable for our security - more complicated. It was a mistake for them to participate as they did. It was a mistake for our presidential candidates to ask them to do so" [@dempsey_military_2016].

Although a professional and non-partisan military is the theoretical goal of the institution, the practical record and results remain far from perfect. From MacArthur's clash with President Truman in the early 1950's, to the so-called "Revolt of the Generals" in 2005-2006, military leaders, even those who are retired, have ventured into political territory on more than one occasion. Nonetheless, academic theorists and military leaders alike have long recognized the importance of professional conduct rooted in non-partisan, apolitical conduct. At its core, such conduct ultimately earns the trust and confidence of the American people.

It must be stressed that the concept of military professionalism does not impose a normative prohibition on the publication of an opinion piece, or the expression of other forms of public dialogue, by a military professional (though there are certainly practical differences between the manner these are employed by those who currently serve and those who are retired). Indeed, Huntington's concept of military professionalism recognizes the unique skills of military officers, some of whom, especially multistar generals and admirals, possess more than 30 and in some cases 40 years of professional experience.

It is reasonable to conclude that a professional opinion piece written by a military officer would take on three very important qualities. First, the military professional would address a relevant military topic. To be clear, multistar generals and admirals often are experienced in the realm of strategy, a word that is often misunderstood and misapplied in discussions of the topic [see for example @strachan_direction_2014]. Without unpacking the concept of strategy further, the relevant point is that military elites, and especially those who have achieved a high rank, have been privy to decisions and experiences occurring near the highest levels of government, and as such, their experiences are not confined to only tactical matters. They are familiar with issues ranging from how wars should be best fought, to future security threats looming on the horizon, to fixing the institutional culture of a military service. Second, from a credibility and expertise perspective, a professional piece would be based on the officer's personal experiences and would be written from an individual's perspective so as not to infer that an argument is being made on behalf of the collective service or element within the military body. Third, an opinion made by a military officer would be free of any hint of partisan sentiment, whether advocacy or denouncement. If one of these three qualities fails to be upheld, there is increasing likelihood that the officer's words can, either intentionally or unintentionally, serve as a weapon, and, like other kinds of weapons, cause great harm. To understand how, we will now turn our attention to understanding the impacts on military professionalism caused by two significant changes that have developed within American Society in the past 50 years. The first is increasing credibility of the military institution, and the second is the introduction of an unparalleled and divisive political polarization. The interaction of these forces is significantly straining the professional conduct of the military institution.

# The Impact of Credibility and Polarization

@@

One of the two significant societal changes that has occurred in America during the past fifty years is the increase in credibility of the military institution. Consider that in the late 1970s, as a likely result of public opinion over the Vietnam War, 54% of Americans indicated that they trusted the military "a great deal or quite a lot," whereas in 2019, 73% of Americas expressed the same sentiments [@gallup_polls_confidence_2007]. Compared to other government institutions, the American public trusts the military to a remarkably high degree. According to a recent Pew Research Center poll, 80% of the American public expects the military to act in the best interests of the American public, compared to business leaders (45%), the media (40%), and elected officials (25%) [@johnson_trust_2018]. It is accurate to say that America trusts its military more than any other entity.

Military leaders are, rightfully so, cognizant of the benefits of holding the public's trust. In the past several years for example, there has been a concerted effort by the services to identify and punish senior leaders who engage in misconduct, to demonstrate to Congress that senior military leaders do not need assistance to handle sexual assault, and to otherwise prove that it can be trusted to accomplish its missions [@us_army_repairing_2014]. Former Chief of Staff of the Army, Retired General Ray Odierno, told fellow Army generals that "of great importance within the Army Profession is the duty of each of us to be a steward of our profession and to maintain its essential and hard won trust with society [Ray Odierno in @us_army_repairing_2014].

@@

Therefore, as the public's trust in the military has grown in the years since Vietnam, so has the military's potential to impact public opinion. This is especially true of military elites, to include retired military officers who previously held substantial rank and positions of authority. An increase in credibility of the institution may not indicate that the military or particular military elites will actually engage in behavior to deliberately impact public opinion, but rather that the potential to do so has increased. The American people tend to listen to those whom they trust.

##Political Polarization Today - Unique in America's History

The second and decisive societal change that has occurred in the past fifty years is the degree to which and character of the political polarization prevalent in the US political landscape today. Although scholars have offered differing explanations for the causes of this polarization, ranging from concerns over race and religion [@abramowitz_great_2018], to the size and role of government [@sides_two_2015], to fierce policy disagreements over policy issues [@mann_its_2012], virtually all scholars agree that polarization in America has become more pronounced in recent decades [@hare_polarization_2014; @bonica_mapping_2014], and that it is, by many measures, worse than it has been during any period since the end of the American Civil War [@hare_polarization_2014]. The polarized political climate today extends to far more issues than previously polarized periods in American history [@layman_party_2002], is marked by stronger negative feelings towards those who hold opposing views [@abramowitz_great_2018], and, at root, involves fundamental issues of both personal and national identity that are not conducive to any reasonable form of compromise [@abramowitz_great_2018; @mohler_jr_gathering_2020]. Through the lens of these characteristics, one can begin to adequately grasp the challenges of polarization to the professional character of the nation's military.

The first way in which today's highly polarizing political climate is different than that of previous periods in American history is its scope. Fewer and fewer issue areas and policy matters are outside of the "spell" of polarization, owing to a phenomenon known as "conflict extension" [@@layman_party_2002]. In previous eras of polarization, a single issue area served as the "battleground" for partisan conflict. Today, though, this is no longer true. Research has consistently shown that the polarizing politics present in the United States today has caused voters who identify as liberals to be more liberal across multiple issue categories, and that the same holds for conservatives [@abramowitz_great_2018]. This includes the full range of social welfare, cultural, and racial issues, all of which are important in the minds of the electorate and partisans [@layman_party_2006]. As a result, the prospect for cooperation on issues that were not politically polarized remained high, whereas today, the prospect for cooperation in general is lower as a result of the shrinking, non-partisan space [@layman_activists_2010].

The second important characteristic of today's polarizing climate involves the strong, negative feelings that each side espouses towards those who hold differing opinions on a variety of issues. Known as "the rise of negative partisanship" [@abramowitz_great_2018], these strongly negative feelings toward those who hold different political opinions is not a phenomenon that has occurred among political elites only; instead, this trend has been shown to have grown among voters since the late 1970s [@abramowitz_great_2018]. Therefore, in addition to there being a diminished opportunity for compromise owing to the proliferation of issues that are polarized, the feelings produced by today's polarized climate are so strong that the other side believes that other is deeply and morally wrong.

The third characteristic of today's polarizing environment is that polarizing issues do not involve matters of simple policy difference, but rather deep and fundamental beliefs about individual and national identities and purposes [@abramowitz_great_2018; @sides_identity_2018; @mohler_jr_gathering_2020]. The election of Donald Trump and his years in office as President have raised the salience on a host of issues, all of which are essential to various parts of the electorate, including attitudes about race, immigration, and religion  [@sides_identity_2018]. Moreover, the American people are aware of the polarized environment, and in a recent survey, indicated that they are concerned the trend is not improving [@schaeffer_far_2020].

In summary, taken together, these three characteristics of contemporary US political polarization are sobering. As a large body of research indicates, it is right to understand the political climate which American finds itself in today not merely as another polarized period, but rather in a tumultuous era that may well challenge the very definition of what America is and what it stands for [@mann_its_2012; @sides_identity_2018; @abramowitz_great_2018; @mohler_jr_gathering_2020]. How the military may be impacted by the forces of increasing institutional credibility and today's political polarization is the subject of the next section.

# When Words Become Weaponized - A Theory of Credibility and Polarization

Figure 1, below, visually describes a general theory of how institutional credibility and political polarization interact to impact military professionalism. From left to right, two significant societal changes (node 'SC') have occurred since the late 1970s, both of which have been described, and as this theory contends, interact, to impact military professionalism. The first societal change, that of institutional credibility, is indicated by the node 'IC', and the second, that of political polarization, is indicated by the node 'PP'. Each of these, in turn, impact military professionalism, indicated by the node 'MP'. Last, though it is outside of the scope of this paper, I contend that changes in military professionalism as a result of the forces of increasing political polarization and institutional credibility, will impact various military outcomes (node 'MO'). These military outcomes vary, and could include things such as military effectiveness, to include the military's ability to fight and win battles [@talmadge_dictators_2015], as well as strategy formulation, to include the military's ability to work with civilians to produce sound strategic advice [@rapp_civil-military_2015]. Indeed, the full degree to which and in what ways polarization will impact the military are not yet fully known.

The theory grounds the two independent variables discussed here, credibility and polarization, as changes that have occurred within American society in the past fifty years. Because these changes stem from changes that have their originate within broader American society, my theory contends that the military as an institution and the military professionals that comprise it are not impervious to their impacts. In other words, my theory assumes that maintaining professional conduct by the military institution requires a deliberate effort and is hard work, despite the fact that behaving in this manner is a long and storied norm. Professional conduct is not a guaranteed nor automatic outcome.

When institutional credibility is high, a demand signal is created by the American public such that future interactions between the military institution and the American public are encouraged and welcomed. From an institutional perspective, the military might respond to the public's demand signal for increased military interaction by increasingly participating in public ceremonies, parades, and other showcases of military prowess (at sporting venues, for instance). From an individual perspective, military elites, especially retired military officers, would be psychologically primed to increasingly dialogue with the American public when military credibility is high. By extension, other outlets such as the press would also be primed to showcase military elites when military credibility runs high, because these elites represent an institution that commands a high degree of public respect.

When military credibility is low, the theory contends the opposite will occur. Since demand for interactions between the American public and the military is muted when institutional credibility for the military is low, the institution will not engage with the public as frequently nor as intimately because the strong demand signal that existed when institutional credibility was high no longer exists. At the individual level, military elites will also not engage as often with the public, nor will the press ask these elites to. In sum, institutional credibility and respect for the military institution impacts the likelihood of the military institution and military elites interacting with the American public. Such interaction can take a variety of forms. When respect for the military is high, the military is likely to interact, speak to, and dialogue with the American public, and when institutional credibility is low, the military is less likely to do so.

```{r, include=FALSE, echo=FALSE}
problem_dag <- dagify(
    MO ~ MP,
    MP ~ PP + IC,
    PP~SC,
    IC~SC,
    coords = tribble(
      ~ name,      ~ x,    ~ y,
      "MO", 1, .5,
      "MP", .7, .5,
      "IC", .3, 1,
      "PP", .3,0,
      "SC", 0, .5
    ))

@@

problem_dag %>% tidy_dagitty() %>% 
  dag_label(labels=c("MO"="Military\n Outcomes", 
                     "MP"="Military\n Professionalism", 
                     "PP"="Political\n Polarization", 
                     "IC"="Institutional\n Credibility", 
                     "SC"="Major Changes\n In American Society")) %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text()+ 
  geom_dag_label_repel(aes(label=label, fill=label), col="white", show.legend=FALSE)+ theme_dag() + labs(title="Figure 1", subtitle="The Impact of Political Polarization and Institutional Credibility on Military Professionalism")

```
The degree and character of political polarization in a society also has the potential to impact the professionalism of the military as an institution. To understand how, recall that the three salient characteristics of the political polarization prevalent in contemporary America include an increase in the number and scope of issues that have become polarized (the phenomenon referred to as "conflict extension" [@layman_party_2002]), an increase in the strength of negative feelings held by supporters of an issue toward the opposition (the rise of "negative partisanship" [@abramowitz_great_2018]), and an increase in the salience of core issues related to individual and national identity [@white_values_2002; @mann_its_2012; @sides_identity_2018]. These three characteristics convey several implications for the military as an institution and its leaders.
